Demo arXiv template

### 2023-01-29 paper/paper.qmd
put up materials for HDI example

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)
library(tidyverse)
library(latex2exp)
library(patchwork)
```

@@

Weighting

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
options(width = 500)
library(tidyverse)
library(latex2exp)
library(patchwork)
library(indri)
library(GGally)
```

@@

Another type of indexes that are commonly used in social science is to rank entities (i.e. countries) on a topic through a range of indicators. This includes university ranking from QS, Times Higher Education; the Sustainable Development Goals (SDG) Index [@SDSN_Bertelsmann_2021] by the Sustainable Development Solutions Network (SDSN), the Human Development Index (HDI) [@UNDP_2021] by United Nations Development Programme (UNDP), etc. These indexes, or rankings, typically have a set of predefined dimensions, or pillars, and indicators are first reduced to these pillars and then reduced to an index. In the following example, we will first show how to compute the Human Development Index (HDI) using the pipeline developed, and then demonstrate how the alternative expressions and parameters can be tested with two functions: `swap_exprs()` and `swap_values()`.

Human Development Index (HDI) measures the development of a country through [...], more than just the economics. It uses three dimensions and four indicators: health (life expectancy at birth), education (expected years of schooling, mean years of schooling), and standard of living (GNI per capita). The technical notes [@UNDP_2021_tech_notes] from United Nation Development Programme have documented the procedures to calculate HDI and we summarise them as below:

  - take log on GNI per capita,
  - rescale the four indicators using mini-max with the provided minimum and maximum, 
  - aggregate the two education indicators using the arithmetic mean
  - aggregate the three dimensions into the index using the geometrical mean

After the initial data import and wrangling, these steps can be calculated in the pipeline,as:

```{r}
dt <- readxl::read_xlsx(path = here::here("data/hdi.xlsx") , skip = 4)
raw <- dt %>%
  janitor::clean_names() %>%
  dplyr::select(-paste0("x", seq(4, 14, 2))) %>%
  dplyr::rename(id = x1, countries = x2) %>%
  dplyr::mutate(across(c(id, human_development_index_hdi:hdi_rank), as.numeric)) %>%
  dplyr::filter(!is.na(id)) %>%
  dplyr::mutate(across(4:7, ~round(.x, digits = 3))) %>% 
  dplyr::select(-c(human_development_index_hdi,
                   gni_per_capita_rank_minus_hdi_rank, 
                   hdi_rank)) 
colnames(raw) <- c("id", "country", "life_exp", "exp_sch", "avg_sch", "gni_pc")

scaling_params <- tibble::tribble(
   ~Dimension, ~Indicator, ~Var,  ~Minimum, ~Maximum,
  "Health",              "Life expectancy at birth (years)",   "life_exp",    "20",          "85",
  "Education",           "Expected years of schooling (years)",  "exp_sch",    "0",          "18",
  "Education",           "Mean years of schooling (years)",   "avg_sch",      "0",          "15",
  "Standard of living",  "GNI per capita (2017 PPP$)",     "gni_pc",       "100",      "75000"
  ) %>%
  mutate(across(contains("mum"), as.numeric),
         across(contains("mum"), ~ifelse(Var == "gni_pc", log10(.x), .x)))
```

```{r echo = TRUE, message=TRUE}
dt <- raw %>% init(id = country, indicators = life_exp:gni_pc)
(res <- dt %>%
  var_trans(gni_pc = log10(gni_pc)) %>% 
  var_trans(.method = rescale_minmax, .vars = life_exp:gni_pc,
            min = scaling_params$Minimum, max = scaling_params$Maximum) %>% 
  dim_red(sch = (exp_sch + avg_sch) / 2) %>%
  dim_red(index = (life_exp * sch * gni_pc)^(1/3)))
```

The output is an `indri` object where a summary of the operations performed have been provided. The dataset contains the ID, country name,  rescaled variables (`life_exp`, `exp_sch`, `avg_sch`, `gni_pc`), the average of two indicators in the education dimension (`sch`), and the index calculated (`index`).

A set of questions that can interest index analysts are the what-ifs: how would the index/ ranking changes if a particular parameter or expression changes in the pipeline. In the context of HDI, the index is calculated as the geometrical mean of the three dimensions, what if an arithmetic mean or different weights is used? The maximum and minimum limit is carefully chosen with justification provided in the technical notes:

  > The low minimum value for gross national income (GNI) per capita, $100, is justified by the considerable amount of unmeasured subsistence and nonmarket production in economies close to the minimum, which is not captured in the official data

  What if this lower minimum is raised or decreased?

The function `swap_exprs()` and `swap_values()` can be used to answer these what-ifs.  The two functions first uses a set of arguments to locate the operations the analysts would like to test, and then accepts a list of expressions or values for testing, and lastly, the initial `indri` (raw data). Another five expressions with different weighting emphasis are tested on calculating the final HDI:

```{r echo = TRUE}
res2 <- res %>%
  swap_exprs(
    .var = index,
    .expr = list(
      index1 = (life_exp + sch + gni_pc)/3,
      index2 = 0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc,
      index3 = 0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc,
      index4 = 0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc,
      # PCA recommended weight
      index5 = 0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc), 
    .raw_data = dt)
```

The result contains another 5 variations of indexes calculated from the expression above. Country ranks can be added using the base function `rank()`. Figure @fig-hdi-expr shows the scatterplot matrix of the country ranking against each other, including the original index (`index0`). From the plot, we learn that the geoemtrical mean (`rank0`), the arithmetic mean (`rank1`), and the PCA recommended weights (`rank5`) give almost similar ranks. This validates the method chosen by the original index. However, when the three dimensions are given different weights (`rank2`, `rank3`, and `rank4`), the country ranking can vary substantially. A cluster of points have been highlighted in the panel `rank3` vs. `rank4`. These countries are ranked highly according to `rank4` but become mediocre when ranked by `index3`. Looking into the three dimensions of these countries in Table @tbl-hdi-expr shows that these countries (Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) have high life expectancy and GNI, but relatively low score on education, hence they are ranked lower when education (`sch`) is given a larger weight in `index3` (`0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc`).

```{r fig-hdi-expr}
#| fig-height: 8
#| fig-width: 12
#| fig-cap: "my plot"
with_rank <- res2$data %>% 
  mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>% 
  rename_with(~gsub(".index", "", .x), contains("rank.index"))

lower_cluster <- with_rank %>% 
  filter(rank4 < 30, rank3 > 50) %>% 
  select(country, life_exp, sch, gni_pc, index3, index4, rank3, rank4)

with_highlight <- with_rank %>% 
  mutate(highlight = ifelse(country %in% lower_cluster$country, TRUE, FALSE))

p1 <- ggpairs(
  with_highlight, columns = 14:19,
  upper = list(continuous = "points", combo = "facethist", 
               discrete = "facetbar", na = "na"),
  lower = NULL, diag = NULL) +
  theme(aspect.ratio = 1)

# https://stackoverflow.com/questions/65091593/how-to-mark-certain-point-in-ggpairs
p1$plots[[23]]$mapping <- 
    `class<-`(c(p1$plots[[23]]$mapping, aes(color = highlight)), "uneval")
p1 <- p1 + scale_color_manual(values = c("black", "red"))
p1
```

```{r echo = FALSE}
#| tbl-cap:  my tables
#| label: tbl-hdi-expr
lower_cluster %>% 
  knitr::kable(digits = 3)
```

```{r echo = TRUE}
res3 <- res %>%
  swap_values(
    .module = var_trans, .step = rescale_minmax, .res = gni_pc,
    .var = min, .values = log10(700),
    .raw_data = dt)
```

### 2023-01-29 paper/paper.tex
put up materials for HDI example

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Weighting

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, interior hidden]}{\end{tcolorbox}}\fi

@@

Another type of indexes that are commonly used in social science is to
rank entities (i.e.~countries) on a topic through a range of indicators.
This includes university ranking from QS, Times Higher Education; the
Sustainable Development Goals (SDG) Index (Network and Stiftung 2021) by
the Sustainable Development Solutions Network (SDSN), the Human
Development Index (HDI) (Programme 2021a) by United Nations Development
Programme (UNDP), etc. These indexes, or rankings, typically have a set
of predefined dimensions, or pillars, and indicators are first reduced
to these pillars and then reduced to an index. In the following example,
we will first show how to compute the Human Development Index (HDI)
using the pipeline developed, and then demonstrate how the alternative
expressions and parameters can be tested with two functions:
\texttt{swap\_exprs()} and \texttt{swap\_values()}.

Human Development Index (HDI) measures the development of a country
through {[}\ldots{]}, more than just the economics. It uses three
dimensions and four indicators: health (life expectancy at birth),
education (expected years of schooling, mean years of schooling), and
standard of living (GNI per capita). The technical notes (Programme
2021b) from United Nation Development Programme have documented the
procedures to calculate HDI and we summarise them as below:

\begin{itemize}
\tightlist
\item
  take log on GNI per capita,
\item
  rescale the four indicators using mini-max with the provided minimum
  and maximum,
\item
  aggregate the two education indicators using the arithmetic mean
\item
  aggregate the three dimensions into the index using the geometrical
  mean
\end{itemize}

After the initial data import and wrangling, these steps can be
calculated in the pipeline,as:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
\NormalTok{(res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log10}\NormalTok{(gni\_pc)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
            \AttributeTok{min =}\NormalTok{ scaling\_params}\SpecialCharTok{$}\NormalTok{Minimum, }\AttributeTok{max =}\NormalTok{ scaling\_params}\SpecialCharTok{$}\NormalTok{Maximum) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Index pipeline:

Summary: 
\end{verbatim}

\begin{verbatim}
var_trans: `log10(gni_pc)` -> gni_pc
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = life_exp, min = 20.000000, max = 85.000000)` -> life_exp
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = exp_sch, min = 0.000000, max = 18.000000)` -> exp_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = avg_sch, min = 0.000000, max = 15.000000)` -> avg_sch
\end{verbatim}

\begin{verbatim}
var_trans: `rescale_minmax(var = gni_pc, min = 2.000000, max = 4.875061)` -> gni_pc
\end{verbatim}

\begin{verbatim}
dim_red: `(exp_sch + avg_sch)/2` -> sch
\end{verbatim}

\begin{verbatim}
dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index
\end{verbatim}

\begin{verbatim}

Data: 
# A tibble: 191 x 8
      id country                life_exp exp_sch avg_sch gni_pc   sch index
   <dbl> <chr>                     <dbl>   <dbl>   <dbl>  <dbl> <dbl> <dbl>
 1     1 Switzerland               0.984   0.917   0.924  0.983 0.920 0.962
 2     2 Norway                    0.973   1.01    0.867  0.978 0.939 0.963
 3     3 Iceland                   0.964   1.06    0.918  0.955 0.991 0.970
 4     4 Hong Kong, China (SAR)    1.01    0.960   0.815  0.973 0.887 0.954
 5     5 Australia                 0.993   1.17    0.848  0.936 1.01  0.979
 6     6 Denmark                   0.944   1.04    0.864  0.967 0.952 0.954
 7     7 Sweden                    0.969   1.08    0.841  0.952 0.960 0.960
 8     8 Ireland                   0.954   1.05    0.772  1.00  0.912 0.955
 9     9 Germany                   0.933   0.945   0.939  0.952 0.942 0.942
10    10 Netherlands               0.949   1.04    0.839  0.956 0.939 0.948
# ... with 181 more rows
\end{verbatim}

The output is an \texttt{indri} object where a summary of the operations
performed have been provided. The dataset contains the ID, country name,
rescaled variables (\texttt{life\_exp}, \texttt{exp\_sch},
\texttt{avg\_sch}, \texttt{gni\_pc}), the average of two indicators in
the education dimension (\texttt{sch}), and the index calculated
(\texttt{index}).

A set of questions that can interest index analysts are the what-ifs:
how would the index/ ranking changes if a particular parameter or
expression changes in the pipeline. In the context of HDI, the index is
calculated as the geometrical mean of the three dimensions, what if an
arithmetic mean or different weights is used? The maximum and minimum
limit is carefully chosen with justification provided in the technical
notes:

\begin{quote}
The low minimum value for gross national income (GNI) per capita, \$100,
is justified by the considerable amount of unmeasured subsistence and
nonmarket production in economies close to the minimum, which is not
captured in the official data
\end{quote}

What if this lower minimum is raised or decreased?

The function \texttt{swap\_exprs()} and \texttt{swap\_values()} can be
used to answer these what-ifs. The two functions first uses a set of
arguments to locate the operations the analysts would like to test, and
then accepts a list of expressions or values for testing, and lastly,
the initial \texttt{indri} (raw data). Another five expressions with
different weighting emphasis are tested on calculating the final HDI:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res2 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{swap\_exprs}\NormalTok{(}
    \AttributeTok{.var =}\NormalTok{ index,}
    \AttributeTok{.expr =} \FunctionTok{list}\NormalTok{(}
      \AttributeTok{index1 =}\NormalTok{ (life\_exp }\SpecialCharTok{+}\NormalTok{ sch }\SpecialCharTok{+}\NormalTok{ gni\_pc)}\SpecialCharTok{/}\DecValTok{3}\NormalTok{,}
      \AttributeTok{index2 =} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
      \AttributeTok{index3 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
      \AttributeTok{index4 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
      \CommentTok{\# PCA recommended weight}
      \AttributeTok{index5 =} \FloatTok{0.569} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.576} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.586} \SpecialCharTok{*}\NormalTok{ gni\_pc), }
    \AttributeTok{.raw\_data =}\NormalTok{ dt)}
\end{Highlighting}
\end{Shaded}

The result contains another 5 variations of indexes calculated from the
expression above. Country ranks can be added using the base function
\texttt{rank()}. Figure Figure~\ref{fig-hdi-expr} shows the scatterplot
matrix of the country ranking against each other, including the original
index (\texttt{index0}). From the plot, we learn that the geoemtrical
mean (\texttt{rank0}), the arithmetic mean (\texttt{rank1}), and the PCA
recommended weights (\texttt{rank5}) give almost similar ranks. This
validates the method chosen by the original index. However, when the
three dimensions are given different weights (\texttt{rank2},
\texttt{rank3}, and \texttt{rank4}), the country ranking can vary
substantially. A cluster of points have been highlighted in the panel
\texttt{rank3} vs.~\texttt{rank4}. These countries are ranked highly
according to \texttt{rank4} but become mediocre when ranked by
\texttt{index3}. Looking into the three dimensions of these countries in
Table Table~\ref{tbl-hdi-expr} shows that these countries (Andorra,
Qatar, San Marino, Kuwait, and Brunei Darussalam) have high life
expectancy and GNI, but relatively low score on education, hence they
are ranked lower when education (\texttt{sch}) is given a larger weight
in \texttt{index3}
(\texttt{0.1\ *\ life\_exp\ +\ 0.1\ *\ sch\ +\ 0.8\ *\ gni\_pc}).

{\centering \includegraphics{paper_files/figure-pdf/fig-hdi-expr-1.pdf}

\caption{\label{fig-hdi-expr}my plot}

\hypertarget{tbl-hdi-expr}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2727}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1364}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1061}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1061}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1061}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}@{}}
\caption{\label{tbl-hdi-expr}my tables}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
country
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
life\_exp
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
sch
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
gni\_pc
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index4
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank4
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
country
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
life\_exp
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
sch
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
gni\_pc
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
index4
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank3
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
rank4
\end{minipage} \\
\midrule()
\endhead
Andorra & 0.929 & 0.721 & 0.942 & 0.764 & 0.919 & 69 & 27 \\
Qatar & 0.912 & 0.684 & 1.023 & 0.741 & 0.978 & 83 & 5 \\
San Marino & 0.937 & 0.701 & 0.947 & 0.749 & 0.921 & 76 & 25 \\
Kuwait & 0.903 & 0.670 & 0.947 & 0.721 & 0.915 & 96 & 29 \\
Brunei Darussalam & 0.841 & 0.694 & 0.977 & 0.737 & 0.935 & 85 & 20 \\
\bottomrule()
\end{longtable}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{res3 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{swap\_values}\NormalTok{(}
    \AttributeTok{.module =}\NormalTok{ var\_trans, }\AttributeTok{.step =}\NormalTok{ rescale\_minmax, }\AttributeTok{.res =}\NormalTok{ gni\_pc,}
    \AttributeTok{.var =}\NormalTok{ min, }\AttributeTok{.values =} \FunctionTok{log10}\NormalTok{(}\DecValTok{700}\NormalTok{),}
    \AttributeTok{.raw\_data =}\NormalTok{ dt)}
\end{Highlighting}
\end{Shaded}

@@

\leavevmode\vadjust pre{\hypertarget{ref-SDSN_Bertelsmann_2021}{}}%
Network, Sustainable Development Solutions, and Bertelsmann Stiftung.
2021. {``SDG Index and Dashboards.''} \url{https://sdgindex.org/}.

\leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021}{}}%
Programme, United Nations Development. 2021a. {``Human Development Index
(HDI).''}
\url{https://hdr.undp.org/en/content/human-development-index-hdi}.

\leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021_tech_notes}{}}%
---------. 2021b. {``Human Development Report 2021-2022: The Next
Frontier: Human Development and the Anthropocene.''}
\url{https://hdr.undp.org/sites/default/files/2021-22_HDR/hdr2021-22_technical_notes.pdf}.
